---
title: "p3_summarise_ssd.R"
author: "Alex Pate"
date: "2024-12-10"
output: word_document
---

## Including Plots

You can also embed plots, for example:

```{r, include=FALSE}
### Set working directory
rm(list=ls())

### Set working directory
setwd("/mnt/bmh01-rds/mrc-multi-outcome/Project_8.2/")

### Source functions
R.func.sources <- list.files("R", full.names = TRUE)
sapply(R.func.sources, source)

### Load functions
source("R/sim_functions.R")

### Call libraries
library(ggplot2)

### Read in data
metadata <- readRDS("data/heuristic_shrinkage_sim_metadata_ssd.rds")
table(metadata$corr)

### Split metadata into boservations both are recorded, and obserations just Riley recorded
metadata.both <- subset(metadata, !is.na(S.pop.mean.riley) & !is.na(S.pop.mean.pavlou))
metadata.all.riley <- subset(metadata, !is.na(S.pop.mean.riley))
metadata.just.riley <- subset(metadata, !is.na(S.pop.mean.riley) & is.na(S.pop.mean.pavlou))

### See how many just riley
sum(is.na(metadata$S.pop.mean.riley) & is.na(metadata$S.pop.mean.pavlou))
sum(!is.na(metadata$S.pop.mean.riley) & is.na(metadata$S.pop.mean.pavlou))
sum(!is.na(metadata$S.pop.mean.riley) & is.na(metadata$S.pop.mean.pavlou))/nrow(metadata)

### Write a function to get the quantiles for a given range of C-statistics
get_quantiles <- function(C1,C2,correlated = TRUE, both = TRUE, magnitude = FALSE){
  
  if (both == TRUE){
    ## Subset
    df <- subset(metadata.both, C.pop > C1 & C.pop <= C2 & corr == correlated)
    
    ## Calculate quantiles
    quantiles.riley <- quantile(as.numeric(df$S.pop.mean.riley), p = c(0.025,0.25,0.5,0.75,0.975))
    quantiles.pavlou <- quantile(as.numeric(df$S.pop.mean.pavlou), p = c(0.025,0.25,0.5,0.75,0.975))
    
    ## Calculate mean/median difference from 0.9
    if (magnitude == FALSE){
      mean.riley <- mean(as.numeric(df$S.pop.mean.riley) - 0.9)
      names(mean.riley) <- "mean"
      median.riley <- median(as.numeric(df$S.pop.mean.riley) - 0.9)
      names(median.riley) <- "median"
      
      mean.pavlou <- median(as.numeric(df$S.pop.mean.pavlou) - 0.9)
      names(mean.pavlou) <- "mean"
      median.pavlou <- median(as.numeric(df$S.pop.mean.pavlou) - 0.9)
      names(median.pavlou) <- "median"
    } else {
      mean.riley <- mean(abs(as.numeric(df$S.pop.mean.riley) - 0.9))
      names(mean.riley) <- "mean"
      median.riley <- median(abs(as.numeric(df$S.pop.mean.riley) - 0.9))
      names(median.riley) <- "median"
      
      mean.pavlou <- median(abs(as.numeric(df$S.pop.mean.pavlou) - 0.9))
      names(mean.pavlou) <- "mean"
      median.pavlou <- median(abs(as.numeric(df$S.pop.mean.pavlou) - 0.9))
      names(median.pavlou) <- "median"
    }

    ## Concatenate
    quantiles.riley <- c(quantiles.riley, mean.riley, median.riley)
    quantiles.pavlou <- c(quantiles.pavlou, mean.pavlou, median.pavlou)
    
  } else {
    ## Subset
    df <- subset(metadata.just.riley, C.pop > C1 & C.pop <= C2 & corr == correlated)
    
    ## Calculate quantiles
    quantiles.riley <- quantile(as.numeric(df$S.pop.mean.riley), p = c(0.025,0.25,0.5,0.75,0.975))
    quantiles.pavlou <- NA
    
    ## Calculate mean/median difference from 0.9
    if (magnitude == FALSE){
      mean.riley <- mean(as.numeric(df$S.pop.mean.riley) - 0.9)
      names(mean.riley) <- "mean"
      median.riley <- median(as.numeric(df$S.pop.mean.riley) - 0.9)
      names(median.riley) <- "median"
      
      mean.pavlou <- NA
      names(mean.pavlou) <- "mean"
      median.pavlou <- NA
      names(median.pavlou) <- "median"
    } else {
      mean.riley <- mean(abs(as.numeric(df$S.pop.mean.riley) - 0.9))
      names(mean.riley) <- "mean"
      median.riley <- median(abs(as.numeric(df$S.pop.mean.riley) - 0.9))
      names(median.riley) <- "median"
      
      mean.pavlou <- NA
      names(mean.pavlou) <- "mean"
      median.pavlou <- NA
      names(median.pavlou) <- "median"
    }

    ## Concatenate
    quantiles.riley <- c(quantiles.riley, mean.riley, median.riley)
    quantiles.pavlou <- c(quantiles.pavlou, mean.pavlou, median.pavlou)
    
  }
  return(list("riley" = quantiles.riley,
              "pavlou" = quantiles.pavlou))
}

### Write a function to get the quantiles for a given range of C-statistics
get_N <- function(C1,C2,correlated = TRUE, both = TRUE){
  if (both == TRUE){
    df <- subset(metadata.both, C.pop > C1 & C.pop <= C2 & corr == correlated)
  } else {
    df <- subset(metadata.just.riley, C.pop > C1 & C.pop <= C2 & corr == correlated)
  }
  n <- nrow(df)
  paste("N=", n)
}

### Put S_VH by quantiles ranges into a list
quantiles.list <- list(c(0.6, 0.65),
                       c(0.65, 0.7),
                       c(0.7, 0.75),
                       c(0.75, 0.8),
                       c(0.8, 0.85),
                       c(0.85, 0.9))

### Print range of C-statistics for the scenarios where Pavlou approach did not converge
metadata.just.riley.corr <- subset(metadata.just.riley, corr == TRUE)
metadata.just.riley.nocorr <- subset(metadata.just.riley, corr == FALSE)
quantile(metadata.just.riley$C.pop, p = c(0, 0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975))
quantile(metadata.just.riley.corr$C.pop, p = c(0, 0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975))
quantile(metadata.just.riley.nocorr$C.pop, p = c(0, 0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975))
```

# Bias

## Get results for correlated data

```{r, echo = FALSE}
S.pop.by.quantiles <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2])})
S.pop.by.quantiles.riley <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], both = FALSE)})
S.pop.all <- get_quantiles(0,1)
S.pop.all.riley <- get_quantiles(0,1, both = FALSE)

### Put into a table
S.pop.by.quantiles <- lapply(S.pop.by.quantiles, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
})
S.pop.all <- rbind(S.pop.all[[1]], S.pop.all[[2]])

### Suppress warnings for scenarios where N_adapted fails, as they give all NA, but we are only interetes in the results for N_original
S.pop.by.quantiles.riley <- suppressWarnings(lapply(S.pop.by.quantiles.riley, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
}))
S.pop.all.riley <- suppressWarnings(rbind(S.pop.all.riley[[1]], S.pop.all.riley[[2]]))
```

Range of $S_{opt}$ for all, then C-statistics of c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all, digits = 3)
knitr::kable(S.pop.by.quantiles[[1]], digits = 3)
knitr::kable(S.pop.by.quantiles[[2]], digits = 3)
knitr::kable(S.pop.by.quantiles[[3]], digits = 3)
knitr::kable(S.pop.by.quantiles[[4]], digits = 3)
knitr::kable(S.pop.by.quantiles[[5]], digits = 3)
knitr::kable(S.pop.by.quantiles[[6]], digits = 3)
```

Get number of observations total, and in each C-statistic group c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Get N values
get_N(0,1)
lapply(quantiles.list, function(x) {get_N(x[1],x[2])})
```

Range of $S_{opt}$ for scenarios where $N_{adapted}$ failed.

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all.riley, digits = 3)

### Get N values
get_N(0,1, both = FALSE)
```

## Get results for uncorrelated data

```{r}
S.pop.by.quantiles <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], correlated = FALSE)})
S.pop.by.quantiles.riley <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], correlated = FALSE, both = FALSE)})
S.pop.all <- get_quantiles(0,1, correlated = FALSE)
S.pop.all.riley <- get_quantiles(0,1, correlated = FALSE, both = FALSE)

### Put into a table
S.pop.by.quantiles <- lapply(S.pop.by.quantiles, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
})
S.pop.all <- rbind(S.pop.all[[1]], S.pop.all[[2]])

### Suppress warnings for scenarios where N_adapted fails, as they give all NA, but we are only interetes in the results for N_original
S.pop.by.quantiles.riley <- suppressWarnings(lapply(S.pop.by.quantiles.riley, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
}))
S.pop.all.riley <- suppressWarnings(rbind(S.pop.all.riley[[1]], S.pop.all.riley[[2]]))

```

Range of $S_{opt}$ for all, then C-statistics of c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all, digits = 3)
knitr::kable(S.pop.by.quantiles[[1]], digits = 3)
knitr::kable(S.pop.by.quantiles[[2]], digits = 3)
knitr::kable(S.pop.by.quantiles[[3]], digits = 3)
knitr::kable(S.pop.by.quantiles[[4]], digits = 3)
knitr::kable(S.pop.by.quantiles[[5]], digits = 3)
knitr::kable(S.pop.by.quantiles[[6]], digits = 3)
```

Get number of observations total, and in each C-statistic group c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Get N values
lapply(quantiles.list, function(x) {get_N(x[1],x[2], correlated = FALSE)})
get_N(0,1, correlated = FALSE)
```

Range of $S_{opt}$ for scenarios where $N_{adapted}$ failed.

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all.riley, digits = 3)


### Get N values
get_N(0,1, correlated = FALSE, both = FALSE)
```

# Magnitude of bias

## Get results for correlated data

```{r, echo = FALSE}
S.pop.by.quantiles <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], magnitude = TRUE)})
S.pop.by.quantiles.riley <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], both = FALSE, magnitude = TRUE)})
S.pop.all <- get_quantiles(0,1, magnitude = TRUE)
S.pop.all.riley <- get_quantiles(0,1, both = FALSE, magnitude = TRUE)

### Put into a table
S.pop.by.quantiles <- lapply(S.pop.by.quantiles, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
})
S.pop.all <- rbind(S.pop.all[[1]], S.pop.all[[2]])

### Suppress warnings for scenarios where N_adapted fails, as they give all NA, but we are only interetes in the results for N_original
S.pop.by.quantiles.riley <- suppressWarnings(lapply(S.pop.by.quantiles.riley, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
}))
S.pop.all.riley <- suppressWarnings(rbind(S.pop.all.riley[[1]], S.pop.all.riley[[2]]))
```

Range of $S_{opt}$ for all, then C-statistics of c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all, digits = 3)
knitr::kable(S.pop.by.quantiles[[1]], digits = 3)
knitr::kable(S.pop.by.quantiles[[2]], digits = 3)
knitr::kable(S.pop.by.quantiles[[3]], digits = 3)
knitr::kable(S.pop.by.quantiles[[4]], digits = 3)
knitr::kable(S.pop.by.quantiles[[5]], digits = 3)
knitr::kable(S.pop.by.quantiles[[6]], digits = 3)
```

Get number of observations total, and in each C-statistic group c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Get N values
get_N(0,1)
lapply(quantiles.list, function(x) {get_N(x[1],x[2])})
```

Range of $S_{opt}$ for scenarios where $N_{adapted}$ failed.

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all.riley, digits = 3)

### Get N values
get_N(0,1, both = FALSE)
```

## Get results for uncorrelated data

```{r}
S.pop.by.quantiles <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], correlated = FALSE, magnitude = TRUE)})
S.pop.by.quantiles.riley <- lapply(quantiles.list, function(x) {get_quantiles(x[1],x[2], correlated = FALSE, both = FALSE, magnitude = TRUE)})
S.pop.all <- get_quantiles(0,1, correlated = FALSE, magnitude = TRUE)
S.pop.all.riley <- get_quantiles(0,1, correlated = FALSE, both = FALSE, magnitude = TRUE)

### Put into a table
S.pop.by.quantiles <- lapply(S.pop.by.quantiles, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
})
S.pop.all <- rbind(S.pop.all[[1]], S.pop.all[[2]])

### Suppress warnings for scenarios where N_adapted fails, as they give all NA, but we are only interetes in the results for N_original
S.pop.by.quantiles.riley <- suppressWarnings(lapply(S.pop.by.quantiles.riley, function(x) {
  out <- rbind(x[[1]], x[[2]])
  rownames(out) <- c("RvS", "Rvs-adj")
  return(out)
}))
S.pop.all.riley <- suppressWarnings(rbind(S.pop.all.riley[[1]], S.pop.all.riley[[2]]))

```

Range of $S_{opt}$ for all, then C-statistics of c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all, digits = 3)
knitr::kable(S.pop.by.quantiles[[1]], digits = 3)
knitr::kable(S.pop.by.quantiles[[2]], digits = 3)
knitr::kable(S.pop.by.quantiles[[3]], digits = 3)
knitr::kable(S.pop.by.quantiles[[4]], digits = 3)
knitr::kable(S.pop.by.quantiles[[5]], digits = 3)
knitr::kable(S.pop.by.quantiles[[6]], digits = 3)
```

Get number of observations total, and in each C-statistic group c(0.6, 0.65), c(0.65, 0.7), c(0.7, 0.75), c(0.75, 0.8), c(0.8, 0.85), and c(0.85, 0.9).

```{r, echo = FALSE}
### Get N values
lapply(quantiles.list, function(x) {get_N(x[1],x[2], correlated = FALSE)})
get_N(0,1, correlated = FALSE)
```

Range of $S_{opt}$ for scenarios where $N_{adapted}$ failed.

```{r, echo = FALSE}
### Print S_pop by quantiles
knitr::kable(S.pop.all.riley, digits = 3)


### Get N values
get_N(0,1, correlated = FALSE, both = FALSE)
```